---
title: "Models"
output: html_document
date: "2023-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(openxlsx)
library(readxl)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(randomForest)
library(glmnet)
library(cowplot)
library(Boruta) #Feature selection
library(boot)
library(xgboost)
library(grid)
library(IRdisplay)
library(cowplot)
library(boot)
library(caret)
library(mboost)
library(ranger)
library(gam)
library(hrbrthemes)
library(cowplot)
library(kableExtra)
```

### Model Fitting and Evaluation

In this section we explain the procedure to tune, fit, evaluate and save the different models created for each country/region. The code organization for this task corresponds to the following structure in our github repository:

- MainFileCountry.Rmd:
- MainFileRegion.Rmd:
- Functions
  - Modeling
    - Boost_gam.R:
    - Boost_glm.R:
    - Lasso.R:
    - RandomForest.R:
    - RFRanger.R:
    - XGBoost.R:
  - Results
    - ConfidentIntervalGeneral.R:
    - Est_Metrics.R:
    - GetResultsGeneral.R:
    - PlotResultsGeneral.R:
- Figures:
- Imputed_Data:



The main algorithm starts with the data in a long format such as the following.

```{r echo=FALSE, message=FALSE}
Imputed_data<- as.data.frame(read.xlsx(
    "../Imputed_data/Final_imputed_data_All.xlsx"))
kable(head(Imputed_data, 10))%>%
  kable_styling(full_width = FALSE)%>%
  kable_styling(font_size = 14)
```

Then, the data is filtered by each country or region and a series of regression models are tested using a training/testing split approach. For each country/region, the training data is formed by all observations before 2015 and th